Attribute fusion in a latent process model for time series of graphs

### 2013-02-27 ssp2011.tex
Fixed some notations issues
Version accepted to TSP

\documentclass[10pt,twocolumn]{IEEEtran}
\usepackage{fontenc}
\pdfminorversion=4
%\usepackage{lmodern}
%\usepackage[utopia]{mathdesign}
\usepackage{graphicx}
%\usepackage{xltxtra}
%\setmainfont[Mapping=tex-text]{Linux Libertine}
\usepackage[cmex10]{amsmath}
\interdisplaylinepenalty=2500
\usepackage{array}
\usepackage{amssymb}
\usepackage{amsthm}
%\usepackage[tight,normalsize,sf,SF]{subfigure}
\ifCLASSOPTIONcompsoc
\usepackage[font=normalsize,labelfont=sf,textfont=sf]{subfig}
\else
\usepackage[font=footnotesize]{subfig}
\fi
\usepackage{subfig}
\usepackage{parskip}
\usepackage{mathrsfs}
\usepackage{bm}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
%\usepackage[colorlinks=true,pagebackref,linkcolor=magenta]{hyperref}
\usepackage[noadjust]{cite}
\renewcommand\arraystretch{1.2}
\newcommand{\argmax}{\operatornamewithlimits{argmax}}
\newcommand{\argmin}{\operatornamewithlimits{argmin}}
\def\mathllap{\mathpalette\mathllapinternal}
\def\mathllapinternal#1#2{%
\llap{$\mathsurround=0pt#1{#2}$}% $
}
\def\clap#1{\hbox to 0pt{\hss#1\hss}}
\def\mathclap{\mathpalette\mathclapinternal}
\def\mathclapinternal#1#2{%
\clap{$\mathsurround=0pt#1{#2}$}%
}
\def\mathrlap{\mathpalette\mathrlapinternal}
\def\mathrlapinternal#1#2{%
\rlap{$\mathsurround=0pt#1{#2}$}% $
}
\bibliographystyle{IEEEtran}
\begin{document}
\title{Attribute fusion in a latent process model for time series of
  graphs}
\author{Minh~Tang, Youngser~Park, Nam~H.~Lee, and Carey~E.~Priebe*%
\thanks{Minh~Tang is with the Department of Applied Mathematics and
  Statistics, Johns Hopkins University, Baltimore, MD 21218
  USA. (email: mtang10@jhu.edu)}%
\thanks{Youngser~Park is with the Center of Imaging Science, Johns
  Hopkins University, Baltimore, MD 21218
  USA. (email:youngser@jhu.edu)}%
\thanks{Nam~H.~Lee is with the Department of Applied Mathematics and
  Statistics, Johns Hopkins University, Baltimore, MD 21218
  USA. (email: nhlee@jhu.edu)}%
\thanks{Carey~E.~Priebe is with the Department of Applied Mathematics and
  Statistics, Johns Hopkins University, Baltimore, MD 21218
  USA. (phone: 410-516-7200; fax:410-516-7459; email: cep@jhu.edu)}}%
%\markboth{IEEE Transactions on Signal Processing}{}
% \begin{IEEEkeywords}
%   Anomaly detection, Attributed Random Graphs, Random Dot
%   Product Graphs.
% \end{IEEEkeywords}}

\maketitle
\begin{abstract}
  Hypothesis testing on time series of attributed graphs has
  applications in diverse areas, e.g., social network analysis
  (wherein vertices represent individual actors or organizations),
  connectome inference (wherein vertices are neurons or brain regions)
  and text processing (wherein vertices represent authors or
  documents).  We consider the problem of anomaly/change point
  detection given the latent process model for time series of graphs
  with categorical attributes on the edges presented in \cite{lee11}.
  Various attributed graph invariants are considered, and their power
  for detection as a function of a linear fusion parameter is
  presented.  Our main result is that inferential performance in
  mathematically tractable first-order and second-order approximation
  models does provide guidance for methodological choices applicable
  to the exact (realistic but intractable) model. Furthermore, to the
  extent that the exact model is realistic, we may tentatively
  conclude that approximation model investigations have some bearing
  on real data applications.
\end{abstract}
\ifCLASSOPTIONpeerreview
\begin{center} \bfseries EDICS Category: SSP-SSAN, SSP-DETC \end{center}
\fi
%\pagebreak
\section{Introduction}
\label{sec:introduction}
The notion of data as a time series is fundamental to signal
processing. However, time-series analysis had been mostly confined to
data in Euclidean spaces, while the representation of data as graphs
with the vertices as the entities and the edges as the relationships
between the entities is now ubiquitous in many application domains,
e.g., social networks (wherein vertices represent individual actors or
organizations), neuroscience (wherein vertices are neurons or brain
regions), and text processing (wherein vertices represent authors or
documents). Furthermore, in many of these application domains, the
edges between the vertices, in addition to representing the presence
of a relationship, can also be use to identify the attributes
associated with that relationship. For example, in social network
graphs where edges represent email communications, the attributes
signify the context of the email messages. Finally, in many of these
application domains, the underlying data changes over time, and thus a
representation of data as a time-series of attributed graphs is
natural. The analysis of time-series of (attributed) graphs thus
presents a natural and challenging extension to the signal processing
community.

We mention here some related work. The notion of signal
processing for graphs appears to be in an emerging
state, with significant research activity in constructing wavelet
representations for graphs, e.g., 
\cite{hammond10:_wavel,r.coifman06:_diffus,crovella03:_graph,narangss:_perfec}. 
The formulation of graph analysis problems in a signal processing framework
is also of potential interest, e.g, \cite{miller10:_towar_euclid}.

One of the main exploitation tasks in time-series analysis is the
problem of anomaly/change-point detection, where anomaly is broadly
interpreted to mean deviation from some ``normal'' pattern and
change-point is the time window at which the anomalous deviation
occurs. In the context of a time-series of graphs, an anomaly can be
for example the presence of a specific subgraph, or frequent occurence
of some subgraphs, or other more vague notions such as existence of a
community structure.

In this paper, we investigate the anomaly/change-point detection
problem for time-series of attributed graphs. In the spirit of
time-series analysis, we approach this problem via test statistics
based on the moving average of some graph invariants. We derive
limiting distributions for these test statistics under a generative
model for the attributed graphs. The experimental evidence indicate that these
test statistics have sufficient power against the null hypothesis for
a general class of change-point problem but that many interesting
questions remain to ponder.

Our paper is structured as follows. We present an overview of the
generative model for time-series of attributed graphs \cite{lee11} in
\S\ref{sec:latent-process-model}. The change-point detection problem
along with the relevant test statistics are described in
\S\ref{sec:change-point-detect}. The distributions of the test
statistics and the power estimates for hypothesis testing using
these test statistics are given in \S\ref{sec:power-estimates}. 
\S\ref{sec:inference-examples} presents a
synthetic data example and an example using the Enron email corpus to
illustrate our methodology.
\section{Latent Process Model}
\label{sec:latent-process-model}
The abundance of graph representations in diverse application domains
leads to a proliferation of random graph models to model the
association among the entities. In addition to the classical
Erd\"{o}s-R\'{e}nyi model \cite{Erdos1959}, there is, the exponential
graph models \cite{holland81}, the stochastic block model
\cite{Holland1983,Wang1987}, the latent position model
\cite{hoff02:_laten}, the dot product model for unattributed graphs
\cite{young07:_random} and the latent process model for attributed
graphs \cite{lee11}. We now describe briefly the latent position model
of \cite{hoff02:_laten} and the dot product model of
\cite{young07:_random} as they are closely related to the latent
process model in \cite{lee11}. The latent process model serves as
the generative model for our attributed graphs and thus plays a
key role in our setup of the anomaly detection problem in 
\S~\ref{sec:change-point-detect}.

The latent space model is one where each vertex is associated with a
latent random vector. There may also be additional covariate
information but those are not relevant to the current exposition. The
vectors are independent and identically distributed and the
probability of an edge between two nodes depends only on their latent
vectors. Conditioned on the latent vectors, the presence of each edge
is an independent Bernoulli trial. For example, if $p_{uv}$ is the
probability of an edge between the vertices $u$ and $v$, then $p_{uv}
= \tfrac{1}{1 + \exp( - \|z_u - z_v\|)}$ where $z_u$ and $z_v$ are the
latent vectors associated with vertex $u$ and $v$, respectively.
Another example of a latent space model is the random dot product
graph model \cite{young07:_random}. Under the random dot product graph
model, the probability of an edge between two vertices is given
by the dot product of their respective latent vectors, i.e., $p_{uv} =
\langle z_u, z_v \rangle$. Parts of the motivation of the random dot
product graph model is that, in a social network with edges
indicating friendships, the components of the vector may be
interpreted as the relative interest of the individual in various
topics. The magnitude of the vector can be interpreted as how
talkative the individual is, with more talkative individuals more
likely to form relationships. Talkative individuals interested in the
same topics are most likely to form relationships while individuals
who do not share interests are unlikely to form relationships.

@@

We now give the necessary details to make the above ideas precise. Let $G =
(V,E)$ be an undirected graph, equipped with an edge-attribution
function $\phi \colon \tbinom{V}{2} \mapsto [K+1] = \{1,2,\dots, K+1\}$ for
some $K$ such that $\phi(e) \leq K$ for $e \in E$ and $\phi(e) \equiv
K+1$ for $e \not \in E$. We refer to $\phi(e)$ as the attribute of
edge $e \in E$, and the pair $(G,\phi)$ as an attributed graph. Let
$\mathscr{S}$ be the unit simplex in $\mathbb{R}^{K}$, i.e.,
\begin{equation}
  \mathscr{S} = \{ \xi \in [0,1]^{K}
  \colon \sum_{k = 1}^{K} \xi_k \leq 1 \}.
\end{equation}
A {\em random dot product space} for attributed graphs with vertices
in $[n]$ ($[n] = \{1,2,\dots,n\}$) and edge attributes $[K]$ ($[K] =
\{1,2,\dots,K\}$) is a pair $(\mathbf{X},G)$ of
random elements such that
\begin{enumerate}
\item $\mathbf{X} = \{X_v\}_{v = 1}^{n}$ is a collection of
  independent $\mathscr{S}$-valued random vectors.
\item $G$ is a random undirected graph with vertex set $[n]$ such
  that the edges of $G$ are conditionally independent given
  $\mathbf{X}$ and that
  \begin{equation}
    \label{eq:1}
    \mathbb{P}(u \sim v \,|\, \mathbf{X}) = \mathbb{P}(u \sim v \, |
    \, X_u,X_v) = \langle x_u, x_v \rangle.
  \end{equation}
 Furthermore, we also have 
  \begin{equation}
    \label{eq:11}
    \mathbb{P}(\phi(\{u,v\}) = k \, | \, u \sim
    v) = \frac{x_{u,k} x_{v,k}}{\langle x_u, x_v \rangle}
  \end{equation}
  i.e., the pair $\{u,v\}$ is an edge in $G$ with attribute $k$ with
  probability $x_{u,k} x_{v,k}$.  
\end{enumerate}
We say that a c\'{a}dl\'{a}g (right continuous with left limit)
process $W \colon [0,\infty) \mapsto [K+1]^{n}$ induces the sequence
of random dot product spaces $\mathscr{V} = (\mathbf{X}(t), G(t))_{t =
1}^{\infty}$ if
\begin{enumerate}
\item Each $(\mathbf{X}(t), G(t))$ is a random dot product space with
  vertices $[n]$ and edge attributes $[K]$. Furthermore, for each $v
  \in [n]$, $k \in [K]$ and $t \in \mathbb{N}$, we have
  \begin{equation}
    \label{eq:54}
  X_{v,k}(t)  = \int_{t - 1}^{t}{ \mathbf{1}\{W_v(\omega) = k\}\, d\omega}
  \end{equation}
  where we denote $W_{v}(\omega)$ as the $v$-th component of $W(\omega)$ and
  $X_{v,k}$ as the $k$-th component of $X_v$.  
\item  For each $t \in \mathbb{N}$, we have
  \begin{equation}
    \label{eq:2}
    \mathbb{P}(G(t) = g \,|\, \mathscr{F}_{\leq t}) = \mathbb{P}(G(t)
    = g \,|\, \mathbf{X}(t))
  \end{equation}
where $\mathscr{F}_{\leq t}$ is the $\sigma$-field generated by $\{W(s)
  \colon s \leq t\}$.
\end{enumerate}
We will call any pair $(\mathscr{V}, W)$ that satisfies the above
properties a random dot process model.

@@

change-point phenomena for $(X(t),G(t))$ at time $t = t^{*}$. We chose

@@

for ease of discussion. Permutation of the vertex labels does not

@@

\pi_1, \mathbf{Q}_0, \mathbf{Q}_1)$ as the change parameters.  Because
$W$ is completely determined by the change parameters, we often choose

@@

dot process models with change-point phenomena. For a random dot

@@

$\mathscr{V}$ with change parameters $(t^{*}, m, \pi_0, \pi_1,

@@

to $\mathscr{V}$. Of particular interests are the following two

@@

itself is a special instance of the mixed membership stochastic
blockmodel \cite{airoldi08:_mixed}. The unattributed kidney-egg model

@@

  Let $\mathscr{V}$ be a rdpm with change parameters $(t^{*}, m,

@@

Suppose that we have a sequence $\{\mathscr{V}^{r} \colon r > 0 \}$ of
rdpm with vertices $[n]$ and attributes $[K]$ where for each $r > 0$,
$\mathscr{V}^{r}$ has change parameters $(t^{*}, m, \pi_0, \pi_1, r
\mathbf{Q}_0, r \mathbf{Q}_1)$. The parameter $r$ can be thought of as
the vertex process rate, i.e., the waiting time decreases
exponentially as $r$ increases. Let us now consider the sequence of
first approximations $\bar{\mathscr{V}}^{r}$ and second approximations
$\widehat{\mathscr{V}}^{r}$ of $\mathscr{V}^{r}$. We note that
$\bar{\mathscr{V}}^{r_1} = \bar{\mathscr{V}}^{r_2}$ for any $r_1, r_2
> 0$. Let us then denote by $\bar{\mathscr{V}}$ the (a.e.) unique
first order approximation of $\mathscr{V}^{r}$ for $r > 0$. If we
denote by $\widehat{\Sigma}_{0}^{(r)}$ and
$\widehat{\Sigma}_{1}^{(r)}$ the matrices $\widehat{\Sigma}_0$ and
$\widehat{\Sigma}_1$ for $\widehat{\mathscr{V}}^{r}$, then
$\widehat{\Sigma}_0^{(r)} = \widehat{\Sigma}_0^{(1)}/r$ and
$\widehat{\Sigma}_{1}^{(r)} = \widehat{\Sigma}_1^{(1)}/r$. Therefore,
as $r \rightarrow \infty$, $\widehat{\Sigma}_0^{(r)} \rightarrow
\bm{0}$ and $\widehat{\Sigma}_1^{(r)} \rightarrow \bm{0}$ and so
$\widehat{\mathscr{V}}^{r} \overset{\mathrm{d}}{\longrightarrow}
\bar{\mathscr{V}}$. This indicates that for sufficiently large $r$,
there is little statistical difference among the
$\bar{\mathscr{V}}$, $\mathscr{V}^{r}$ and $\widehat{\mathscr{V}}^{r}$
(\cite[Theorem 2]{lee11}). The behavior of
$\widehat{\mathscr{V}}^{r}$ for ``small'' $r$ is less
clear. $\bar{\mathscr{V}}$ will thus serve as the basis for our
subsequent analysis on graph invariants for attributed random
graphs.

>>>>

\documentclass[10pt,draftclsnofoot,onecolumn]{IEEEtran}
\usepackage{fontenc}
\pdfminorversion=4
%\usepackage{lmodern}
%\usepackage[utopia]{mathdesign}
\usepackage{graphicx}
%\usepackage{xltxtra}
%\setmainfont[Mapping=tex-text]{Linux Libertine}
\usepackage[cmex10]{amsmath}
\interdisplaylinepenalty=2500
\usepackage{array}
\usepackage{amssymb}
\usepackage{amsthm}
%\usepackage[tight,normalsize,sf,SF]{subfigure}
\ifCLASSOPTIONcompsoc
\usepackage[font=normalsize,labelfont=sf,textfont=sf]{subfig}
\else
\usepackage[font=footnotesize]{subfig}
\fi
\usepackage{subfig}
\usepackage{parskip}
\usepackage{mathrsfs}
\usepackage{bm}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
%\usepackage[colorlinks=true,pagebackref,linkcolor=magenta]{hyperref}
\usepackage[noadjust]{cite}
\renewcommand\arraystretch{1.2}
\newcommand{\argmax}{\operatornamewithlimits{argmax}}
\newcommand{\argmin}{\operatornamewithlimits{argmin}}
\def\mathllap{\mathpalette\mathllapinternal}
\def\mathllapinternal#1#2{%
\llap{$\mathsurround=0pt#1{#2}$}% $
}
\def\clap#1{\hbox to 0pt{\hss#1\hss}}
\def\mathclap{\mathpalette\mathclapinternal}
\def\mathclapinternal#1#2{%
\clap{$\mathsurround=0pt#1{#2}$}%
}
\def\mathrlap{\mathpalette\mathrlapinternal}
\def\mathrlapinternal#1#2{%
\rlap{$\mathsurround=0pt#1{#2}$}% $
}
\bibliographystyle{IEEEtran}
\begin{document}
\title{Attribute fusion in a latent process model for time series of
  graphs}
\author{Minh~Tang, Youngser~Park, Nam~H.~Lee, and Carey~E.~Priebe*%
  \thanks{Copyright (c) 2012 IEEE. Personal use of this material is permitted. However, permission to use this material for any other purposes must be obtained from the IEEE by sending a request to pubs-permissions@ieee.org.}
\thanks{Minh~Tang is with the Department of Applied Mathematics and
  Statistics, Johns Hopkins University, Baltimore, MD 21218
  USA. (email: mtang10@jhu.edu)}%
\thanks{Youngser~Park is with the Center of Imaging Science, Johns
  Hopkins University, Baltimore, MD 21218
  USA. (email:youngser@jhu.edu)}%
\thanks{Nam~H.~Lee is with the Department of Applied Mathematics and
  Statistics, Johns Hopkins University, Baltimore, MD 21218
  USA. (email: nhlee@jhu.edu)}%
\thanks{Carey~E.~Priebe is with the Department of Applied Mathematics and
  Statistics, Johns Hopkins University, Baltimore, MD 21218
  USA. (phone: 410-516-7200; fax:410-516-7459; email: cep@jhu.edu)}}%
\markboth{IEEE Transactions on Signal Processing}{}
\maketitle
\begin{abstract}
  Hypothesis testing on time series of attributed graphs has
  applications in diverse areas, e.g., social network analysis
  (wherein vertices represent individual actors or organizations),
  connectome inference (wherein vertices are neurons or brain regions)
  and text processing (wherein vertices represent authors or
  documents).  We consider the problem of anomaly/change point
  detection given the latent process model for time series of graphs
  with categorical attributes on the edges presented in \cite{lee11}.
  Various attributed graph invariants are considered, and their power
  for detection as a function of a linear fusion parameter is
  presented.  Our main result is that inferential performance in
  mathematically tractable first-order and second-order approximation
  models does provide guidance for methodological choices applicable
  to the exact (realistic but intractable) model. Furthermore, to the
  extent that the exact model is realistic, we may tentatively
  conclude that approximation model investigations have some bearing
  on real data applications.
\end{abstract}
 \begin{IEEEkeywords}
   Anomaly detection, Attributed Random Graphs, Random Dot
   Product Graphs.
 \end{IEEEkeywords}
%\begin{center} \bfseries EDICS Category: SSP-SSAN, SSP-DETC \end{center}
%\pagebreak
\section{Introduction}
\label{sec:introduction}
The notion of data as a time series is fundamental to signal
processing. However, time-series analysis had been mostly confined to
data in Euclidean spaces, while the representation of data as graphs
with the vertices as the entities and the edges as the relationships
between the entities is now ubiquitous in many application domains,
e.g., social networks (wherein vertices represent individual actors or
organizations), neuroscience (wherein vertices are neurons or brain
regions), and text processing (wherein vertices represent authors or
documents). Furthermore, in many of these application domains, the
edges between the vertices, in addition to representing the presence
of a relationship, can also be used to identify the attributes
associated with that relationship. For example, in social network
graphs where edges represent email communications, the attributes
signify the context of the email messages. Finally, in many of these
application domains, the underlying data changes over time, and thus a
representation of data as a time-series of attributed graphs is
natural. The analysis of time-series of (attributed) graphs thus
presents a natural and challenging extension to the signal processing
community.

One of the main exploitation tasks in time-series analysis is the
problem of anomaly/change-point detection, where anomaly is broadly
interpreted to mean deviation from some ``normal'' pattern and
change-point is the time window at which the anomalous deviation
occurs. In the context of a time-series of graphs, an anomaly can be
for example the presence of a specific subgraph, or frequent occurrences
of some subgraphs, or other more vague notions such as existence of a
community structure.

The notion of signal processing for graphs appears to be in an
emerging state, with significant research activity in constructing
wavelet representations for graphs, e.g.,
\cite{hammond10:_wavel,r.coifman06:_diffus,crovella03:_graph,narangss:_perfec}.
The formulation of graph analysis problems in a signal processing
framework is also of potential interest, e.g,
\cite{miller10:_towar_euclid}, wherein a subgraph detection problem is
presented, test statistics are defined, noise and signal models are
introduced, and ROC analysis is performed. The construction of
wavelets representations on graphs is analogous to the classical
wavelets constructions, but adapted to use the topology of the graphs
instead of the usual Euclidean space topology. For example, the
wavelet representations of \cite{hammond10:_wavel} and
\cite{r.coifman06:_diffus} are constructed using the combinatorial and
normalized Laplacian matrices of the graphs, respectively, while
\cite{crovella03:_graph} defined wavelets via the $n$-hop
neighborhoods. The wavelet construction in \cite{crovella03:_graph}
and its use in {\em spatial} traffic analysis (emphasis in 
\cite{crovella03:_graph}) is similar to the use of locality/scan
statistics for anomaly detection in
\cite{priebe05:_scan_statis_enron_graph} and the current
paper. Indeed, the locality statistic at a vertex $v$ is the size of an
$n$-hop neighborhood starting at $v$ defined via shortest path
distance and the scan statistic is the maximum of the locality
statistics over all the $n$-hop neighborhoods. Traffic analysis by
tracking the coefficients of the wavelet representations is similar to
anomaly detection by tracking the values of the scan statistics.

In this paper, we investigate the anomaly/change-point detection
problem for time-series of attributed graphs. In the spirit of
time-series analysis, we approach this problem via test statistics
based on the moving average of some graph invariants. We derive
limiting distributions for these test statistics under a generative
model for the attributed graphs. The experimental evidences indicate that these
test statistics have sufficient power against the null hypothesis for
a general class of change-point problem but that many interesting
questions remain to ponder.

The current paper builds upon ideas previously presented in
\cite{lee11} and \cite{priebe11:_attrib}, among others. The generative
model for time-series of attributed graphs used in this paper is from
\cite{lee11}. Hypothesis testing for time-series of graphs and
attribute fusion using a linear fusion parameter $\lambda$ were
considered in \cite{lee11,priebe11:_attrib}; limiting distribution for
the size graph invariant was given in \cite{lee11}. A version of the
experimental results for simulated data included in this paper was first
presented in \cite{priebe11:_attrib}.

The paper is structured as follows. We present an overview of the
generative model for time-series of attributed graphs \cite{lee11} in
\S~\ref{sec:latent-process-model}. A change-point detection problem
along with some relevant test statistics are described in
\S~\ref{sec:change-point-detect}. The distributions of these test
statistics and the power estimates for hypothesis testing using
these test statistics are given in \S~\ref{sec:power-estimates}. 
\S~\ref{sec:inference-examples} presents a
synthetic data example and an example using the Enron email corpus to
illustrate our methodology.

\section{Latent Process Model}
\label{sec:latent-process-model}
The abundance of graph representations in diverse application domains
leads to a proliferation of random graph models to model the
association among the entities. In addition to the classical
Erd\"{o}s-R\'{e}nyi model \cite{Erdos1959}, there is the exponential
graph models \cite{holland81}, the stochastic block model
\cite{Holland1983,Wang1987}, the latent position model
\cite{hoff02:_laten}, the dot product model for unattributed graphs
\cite{young07:_random} and the latent process model for attributed
graphs \cite{lee11}. We now describe briefly the latent position model
of \cite{hoff02:_laten} and the dot product model of
\cite{young07:_random} as they are closely related to the latent
process model in \cite{lee11}. The latent process model serves as
the generative model for our attributed graphs and thus plays a
key role in our setup of the anomaly detection problem in 
\S~\ref{sec:change-point-detect}.

The latent space model is one where each vertex is associated with a
latent random vector. There may also be additional covariates
information but those are not relevant to the current exposition. The
vectors are independent and identically distributed and the
probability of an edge between two nodes depends only on their latent
vectors. Conditioned on the latent vectors, the presence of each edge
is an independent Bernoulli trial. For example, if $p_{uv}$ is the
probability of an edge between the vertices $u$ and $v$, then $p_{uv}
= \tfrac{1}{1 + \exp( - \|z_u - z_v\|)}$ where $z_u$ and $z_v$ are the
latent vectors associated with vertex $u$ and $v$, respectively.
Another example of a latent space model is the random dot product
graph model \cite{young07:_random}. Under the random dot product graph
model, the probability of an edge between two vertices is given
by the dot product of their respective latent vectors, i.e., $p_{uv} =
\langle z_u, z_v \rangle$. A motivation behind the random dot
product graph model is that, in a social network with edges
indicating friendships, the components of the vector may be
interpreted as the relative interest of the individual in various
topics. The magnitude of the vector can be interpreted as how
talkative the individual is, with more talkative individuals more
likely to form relationships. Talkative individuals interested in the
same topics are most likely to form relationships while individuals
who do not share interest are unlikely to form relationships.

@@

We now give the necessary details to make the above ideas precise. For
any positive integer $n$, we shall denote by $[n]$ the set of numbers
$\{1,2,\dots,n\}$. Let $G = (V,E)$ be an undirected graph, equipped
with an edge-attribution function $\phi \colon \tbinom{V}{2} \mapsto
[K+1]$ for some positive integer $K$. We shall require that $\phi$
satisfies $\phi(e) \leq K$ for $e \in E$ and $\phi(e) \equiv K+1$ for
$e \not \in E$. We refer to $\phi(e)$ as the attribute of edge $e \in
E$, and the pair $(G,\phi)$ as an attributed graph. Let $\mathscr{S}$
be the unit simplex in $\mathbb{R}^{K}$, i.e.,
\begin{equation}
  \mathscr{S} = \{ \xi \in [0,1]^{K}
  \colon \sum_{k = 1}^{K} \xi_k \leq 1 \}.
\end{equation}

A {\em random dot product space} for attributed graphs with vertices
in $[n]$ and edge attributes $[K]$ is a pair $(\mathbf{X},G)$ of
random elements such that
\begin{enumerate}
\item $\mathbf{X} = \{X_v\}_{v = 1}^{n}$ is a collection of
  independent $\mathscr{S}$-valued random vectors.
\item $G$ is a random undirected graph with vertex set $[n]$ such
  that the edges of $G$ are conditionally independent given
  $\mathbf{X}$ and that
  \begin{equation}
    \label{eq:1}
    \mathbb{P}(u \sim v \,|\, \mathbf{X}) = \mathbb{P}(u \sim v \, |
    \, X_u,X_v) = \langle X_u, X_v \rangle.
  \end{equation}
 Furthermore, we also have 
  \begin{equation}
    \label{eq:11}
    \mathbb{P}(\phi(\{u,v\}) = k \, | \, u \sim
    v) = \frac{X_{u,k} X_{v,k}}{\langle X_u, X_v \rangle}
  \end{equation}
  i.e., the pair $\{u,v\}$ is an edge in $G$ having attribute $k$ with
  probability $X_{u,k} X_{v,k}$.  
\end{enumerate}
We say that a c\'{a}dl\'{a}g (right continuous with left limit)
process $W \colon [0,\infty) \mapsto [K+1]^{n}$ induces the sequence
of random dot product spaces $\mathscr{V} = (\mathbf{X}(t), G(t))_{t =
1}^{\infty}$ if
\begin{enumerate}
\item Each $(\mathbf{X}(t), G(t))$ is a random dot product space with
  vertices $[n]$ and edge attributes $[K]$. Furthermore, for each $v
  \in [n]$, $k \in [K]$ and $t \in \mathbb{N}$, we have
  \begin{equation}
    \label{eq:54}
  X_{v,k}(t)  = \int_{t - 1}^{t}{ \mathbf{1}\{W_v(\omega) = k\}\, d\omega}
  \end{equation}
  where we denote $W_{v}(\omega)$ as the $v$-th component of $W(\omega)$ and
  $X_{v,k}$ as the $k$-th component of $X_v$.  
\item  For each $t \in \mathbb{N}$, we have
  \begin{equation}
    \label{eq:2}
    \mathbb{P}(G(t) = g \,|\, \mathscr{F}_{\leq t}) = \mathbb{P}(G(t)
    = g \,|\, \mathbf{X}(t))
  \end{equation}
where $\mathscr{F}_{\leq t}$ is the $\sigma$-field generated by $\{W(s)
  \colon s \leq t\}$.
\end{enumerate}
We will call any pair $(\mathscr{V}, W)$ that satisfies the above
properties a random dot process model.

@@

change-point phenomenon for $(X(t),G(t))$ at time $t = t^{*}$. We chose

@@

for ease of discussion. Permutations of the vertex labels do not

@@

\pi_1, \mathbf{Q}_0, \mathbf{Q}_1)$ as the change-parameters.  Because
$W$ is completely determined by the change-parameters, we often choose

@@

dot process models with change-point phenomenon. For a random dot

@@

$\mathscr{V}$ with change-parameters $(t^{*}, m, \pi_0, \pi_1,

@@

to $\mathscr{V}$. Of particular interest are the following two

@@

itself is a special instance of the stochastic
block model \cite{Holland1983}. The unattributed kidney-egg model

@@

  Let $\mathscr{V}$ be a rdpm with change-parameters $(t^{*}, m,

@@

Suppose that we have a sequence $\{\mathscr{V}^{r} \colon r > 0 \}$ of
rdpm with vertices $[n]$ and attributes $[K]$ where for each $r > 0$,
$\mathscr{V}^{r}$ has change-parameters $(t^{*}, m, \pi_0, \pi_1, r
\mathbf{Q}_0, r \mathbf{Q}_1)$. The parameter $r$ can be thought of as
the vertex process rate, i.e., the waiting time decreases
exponentially as $r$ increases. Let us now consider the sequence of
first approximations $\bar{\mathscr{V}}^{r}$ and second approximations
$\widehat{\mathscr{V}}^{r}$ of $\mathscr{V}^{r}$. We note that
$\bar{\mathscr{V}}^{r_1} = \bar{\mathscr{V}}^{r_2}$ for any $r_1, r_2
> 0$. Let us then denote by $\bar{\mathscr{V}}$ the (a.e.) unique
first order approximation of $\mathscr{V}^{r}$ for $r > 0$. If we
denote by $\widehat{\Sigma}_{0}^{(r)}$ and
$\widehat{\Sigma}_{1}^{(r)}$ the matrices $\widehat{\Sigma}_0$ and
$\widehat{\Sigma}_1$ for $\widehat{\mathscr{V}}^{r}$, then
$\widehat{\Sigma}_0^{(r)} = \widehat{\Sigma}_0^{(1)}/r$ and
$\widehat{\Sigma}_{1}^{(r)} = \widehat{\Sigma}_1^{(1)}/r$. Therefore,
as $r \rightarrow \infty$, $\widehat{\Sigma}_0^{(r)} \rightarrow
\bm{0}$ and $\widehat{\Sigma}_1^{(r)} \rightarrow \bm{0}$ and so
$\widehat{\mathscr{V}}^{r} \overset{\mathrm{d}}{\longrightarrow}
\bar{\mathscr{V}}$. This indicates that for sufficiently large $r$,
there is little statistical difference among the
$\bar{\mathscr{V}}$, $\mathscr{V}^{r}$ and $\widehat{\mathscr{V}}^{r}$
(\cite[Theorem 2]{lee11}). The behavior of
$\widehat{\mathscr{V}}^{r}$ for ``small'' $r$ is less
clear. $\bar{\mathscr{V}}$ will thus serve as the basis for our
subsequent analysis on graph invariants for attributed random
graphs.

A comment should also be made about the utility of the exact
generative model. It is computationally intractable to estimate the
parameters of the underlying continuous time Markov chains, i.e. the
$\{\mathbf{Q}^{(v)}(t)\}$, given a snapshot of the graphs. Therefore,
the main interest of \cite{lee11}, and consequently of this paper,
lies not in the exact generative model but in its first and
second-order approximation. We will not be concerned with the
estimation of the latent position vectors for the first and
second-order approximation in this paper, but we note that they can be
done via spectral embedding techniques a la \cite{STFP-2011}. We also
note that more refined models, where the messaging events between each pairs of
vertices are modeled as point process,
e.g. \cite{Heard2010,PerryWolfe2010}, can also be formulated. The
snapshot of the graphs is then a binning of these point processes and
parameter estimation is fundamental to
these approaches.
\section{Change-point detection}
\label{sec:change-point-detect}
\begin{figure}[htbp]
  \centering
  \includegraphics[width=8cm]{graphics/Fig1-SSP2011.pdf}
  \caption{Notional depiction of the problem of change-point detection
    in a time-series of graphs}
  \label{fig:notional_change_point}
\end{figure}
Let $\mathscr{V}$ be a random dot process model with change-parameters
$(t^{*}, m, \pi_0, \pi_1, \mathbf{Q}_0, \mathbf{Q}_1)$. The
change-parameters encapsulate a notion of chatter anomalies, i.e., a
subset of vertices of $\mathscr{V}$ with altered communication
behavior in an otherwise stationary setting as depicted in
Fig.~\ref{fig:notional_change_point}. We are interested in the problem
of testing, for a $t \in \mathbb{N}$, the hypotheses that $t$ is the
change-point of $\mathscr{V}$, namely
\begin{gather*}
  \mathscr{H}_0 \colon t^{*} > t \\
  \mathscr{H}_A \colon t^{*} = t
\end{gather*}
This will be done using the notion of fusion of attributed graph
invariants. The particular invariants of interest are the size
$\mathcal{E}$, number of triangles $\tau$, scan $\Psi$, and max degree
$\Delta$. The scan statistics $\Psi$ are introduced and applied to the
problem of detecting chatter anomalies in
\cite{priebe05:_scan_statis_enron_graph}. The number of triangles is
the simplest graph invariant based on clique counts and one can define
other test statistics for $k$-clique counts with $k \geq 4$.  The use
of various graph invariants as test statistics was considered in
\cite{pao11:_statis_infer_random_graph} and it was shown there, via
Monte Carlo analysis, that no single invariant is uniformly most
powerful.
